---
title: "Multiparameter models"
author: "Math 315: Bayesian Statistics"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "hygge"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: middle, clear

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(dev = 'svg')
library(gridExtra)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(rethinking)
Howell1 <- readr::read_delim("https://raw.githubusercontent.com/rmcelreath/rethinking/master/data/Howell1.csv", delim = ";")
adults <- Howell1 %>% filter(age >= 18)
```

## 1. Introduction to multiparameter models

---

.left-column[
## Example
]

.right-column[

.large[
- Interest is in analyzing the average height of an adult for the Dobe area !Kung San, a foraging population
]

```{r echo=FALSE, message=FALSE, fig.height = 3.5, fig.width = 5, fig.align='center'}
ggplot(adults) +
  geom_histogram(aes(x = height), bins = 20) +
  theme_light()
```

]

---
class: middle, inverse

# Noninformative analysis

---

# Sampling from the joint posterior

.large[
**Target:** $p(\mu |y_1, \ldots, y_m)$

**1.** Sample $\sigma^2$ from $\sigma^2 | y_1, \ldots, y_m \sim \text{InverseGamma}\left( \frac{n-1}{2},\ \frac{(n-1)s^2}{2} \right)$

```{r cache=TRUE}
n <- nrow(adults)
s2 <- var(adults$height)
sigma2s <- 1 / rgamma(1e4, (n-1)/2, (n-1) * s2 / 2) 
```

**2.** Sample $\mu$ from $\mu | \sigma^2, y_1, \ldots, y_m \sim \mathcal{N}\left( \bar{y},\ \frac{\sigma^2}{n} \right)$

```{r cache=TRUE}
ybar <- mean(adults$height)
mus <- rnorm(1e4, ybar, sqrt(sigma2s / n))
```
]


---
class: clear

```{r fig.height=4, fig.width=4, cache=TRUE, fig.align='center', out.width="60%"}
post_samples <- data.frame(mu = mus, sigma2 = sigma2s)
ggplot(data = post_samples) +
  geom_point(mapping = aes(x = mu, y = sigma2), alpha = 0.1) +
  theme_minimal() +
  labs(x = expression(mu), y = expression(sigma^2))
```

---
class: clear

```{r fig.height=4, fig.width=4, cache=TRUE, fig.align='center', out.width="60%"}
ggplot(data = post_samples) +
  geom_point(mapping = aes(x = mu, y = sigma2), alpha = 0.1) +
  geom_density_2d(mapping = aes(x = mu, y = sigma2), color = "orange", size = 0.8) +
  theme_minimal() +
  labs(x = expression(mu), y = expression(sigma^2))
```

---
class: clear

```{r fig.height=4, fig.width=4, cache=TRUE, fig.align='center', out.width="60%"}
ggplot(data = post_samples) +
  geom_density_2d(mapping = aes(x = mu, y = sigma2), color = "orange") +
  theme_minimal() +
  labs(x = expression(mu), y = expression(sigma^2))
```

---
class: clear

```{r echo=FALSE, fig.height=5.5, fig.width=5.5, cache=TRUE, fig.align='center', out.width = "80%"}
p <- ggplot(data = post_samples) +
  geom_point(mapping = aes(x = mu, y = sigma2), alpha = 0.1) +
  geom_density_2d(mapping = aes(x = mu, y = sigma2), color = "orange", size = 0.8) +
  theme_minimal() +
  labs(x = expression(mu), y = expression(sigma^2))

ggExtra::ggMarginal(
  p,
  type = 'density',
  margins = 'both',
  size = 5,
  fill = 'darkorange',
  alpha = 0.6
)
```

---

## Marginal posteriors

.pull-left[

$\boldsymbol \mu$

```{r echo=FALSE, fig.height=2, fig.width=4}
ggplot(data = post_samples) +
  geom_density(mapping = aes(x = mu), color = NA, fill = "darkorange", alpha = 0.5) +
  theme_minimal() +
  labs(x = expression(mu))
```

```{r}
quantile(mus, probs = c(0.05, 0.9))
```

]

.pull-right[

$\boldsymbol \sigma$

```{r echo=FALSE, fig.height=2, fig.width=4}
ggplot(data = post_samples) +
  geom_density(mapping = aes(x = sigma2), color = NA, fill = "darkorange", alpha = 0.5) +
  theme_minimal() +
  labs(x = expression(sigma^2))
```

```{r}
quantile(sigma2s, probs = c(0.05, 0.9))
```

```{r}
quantile(sqrt(sigma2s), probs = c(0.05, 0.9))
```
]

---
class: middle, inverse

# Informative analysis

---

# McElreath's analysis

.pull-left[

### Model

$y_i \sim \mathcal{N}(\mu, \sigma)$

$\mu \sim \mathcal{N}(178,\ 20)$

$\sigma \sim \text{Unif}(0, 50)$

]

.pull-right[

```{r echo=FALSE, fig.height=2, fig.width=4, message=FALSE, warning=FALSE}
prior.pred <- tibble(
  mu =  rnorm( 1e4 , 178 , 20 ),
  sigma = runif( 1e4 , 0 , 50 ),
  height = rnorm( 1e4 , mu , sigma )
)

ggplot(data = prior.pred) +
  geom_histogram(mapping = aes(x = height), fill = "steelblue", bins = 50, alpha = 0.7) +
  labs(x = "Height (in cm)",
       title = "Prior predictive distribution") +
  theme_minimal()
```

.code70[
```{r echo=FALSE}
round(summary(prior.pred$height), 1)
```
]

]

---

# Joint posterior via grid approximation

```{r cache=TRUE}
# Create grid over the coordinate plane
param_grid <- expand.grid(
  mu = seq(from = 118, to = 238, length.out = 1000), 
  sigma = seq(from = 0, to = 50, length.out = 1000)
)

# Calculate joint log prior for each point on grid
logprior <- dnorm(param_grid[, "mu"], 178, 20, log = TRUE) +
      dunif(param_grid[, "sigma"], 0, 50, log = TRUE)

# Calculate log likelihood for each point on grid
ll <- numeric(length = nrow(param_grid))
for(i in 1:nrow(param_grid)) {
  ll[i] <- sum(dnorm(adults$height, mean = param_grid[i, "mu"], 
                     sd = param_grid[i, "sigma"], log = TRUE))
}

# Calculate log posterior, then exponentiate
logposterior <- logprior + ll
unstd_posterior <- exp(logposterior - max(logposterior)) # numeric stability
posterior <- unstd_posterior / sum(unstd_posterior)
```

---

# Joint posterior via grid approximation

```{r}
posterior_draws <- dplyr::sample_n(
  param_grid, size = 1e4, replace = TRUE, weight = posterior
)
```

```{r echo=FALSE, fig.height=5, fig.width=5, cache=TRUE, fig.align='center', out.width = "55%"}
ggplot(data = posterior_draws, aes(x = mu, y = sigma)) +
  geom_point(alpha = 0.2) +
  geom_density2d(color = "orange", size = 0.8) +
  theme_minimal() +
  labs(x = expression(mu),
       y = expression(sigma))
```

---

# Marginal posteriors via grid approximation

.pull-left[

### $\mu$

```{r echo=FALSE, fig.height=2, fig.width=4}
ggplot(data = posterior_draws) +
  geom_density(mapping = aes(x = mu), color = NA, fill = "darkorange", alpha = 0.5) +
  theme_minimal() +
  labs(x = expression(mu))
```

```{r}
quantile(posterior_draws$mu, probs = c(0.05, 0.95))
```

]

.pull-right[

### $\sigma$

```{r echo=FALSE, fig.height=2, fig.width=4}
ggplot(data = posterior_draws) +
  geom_density(mapping = aes(x = sigma), color = NA, fill = "darkorange", alpha = 0.5) +
  theme_minimal() +
  labs(x = expression(sigma^2))
```

```{r}
quantile(posterior_draws$sigma, probs = c(0.05, 0.95))
```

<br>

<br>

]


---

# Vectorization in R

The grid approximation on a 1000 $\times$ 1000 grid took about 30 seconds

Vectorizing the log-likelihood calculation improves speed to 14 seconds

```{r cache=TRUE}
# vectorized log likelihood function
llnorm <- function(x, mu, sigma) { 
  sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}

# use all of the x vector for each mu-sigma combo
llnorm <- Vectorize(llnorm, vectorize.args = c("mu", "sigma"))

# Calculate log likelihood for each point on grid
ll <- llnorm(x = adults$height, mu = param_grid$mu, sigma = param_grid$sigma)

# Everything else is the same...
```
