---
title: "Generalized linear models"
author: "Math 315: Bayesian Statistics"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "hygge"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(dev = 'svg')
library(rjags)
library(coda)
library(ggplot2)
library(ggthemes)
library(rjags)
library(coda)
library(broom)
```

# Arthritis clinical trial

.large[
A double-blind clinical trial investigated a new treatment for rheumatoid arthritis 

We'll focus on a subset of the variables:

Variable | Description
----------|---------------------
`Better`    | whether the drug improved symptoms <br> 1 = yes, 0 = no
`Treatment` | Placebo or Treated
`Sex`       | Male or Female
`Age`       | Age in years
]

---
class: inverse

# Your turn

.Large[
1. What distribution should we use for the response variable, `Better`?

2. How can you transform a linear combination of the predictor to take on values between 0 and 1?

$$\eta_i = \beta_0 + \beta_1 {\tt Treatment} + \beta_2 {\tt Sex} + \beta_3 {\tt Age}$$
]

---

# Data preparation

.large[Load and manipulate data]

```{r}
arthritis <- read.csv("http://aloy.rbind.io/data/arthritis.csv")
```

.large[JAGS needs numeric variables, so convert factors to indicators]

```{r}
X <- cbind(intercept = 1, arthritis[,2:4])
X$Treatment <- as.numeric(X$Treatment == "Treated")
X$Sex <- as.numeric(X$Sex == "Male")

dat_list <- list(Y = arthritis$Better, X = X, 
                 n = nrow(arthritis), p = ncol(X))
```

---

# Model fitting

```{r cache=TRUE, message=FALSE}
# Logistic regression specification
model_string <- textConnection("model{
for(i in 1:n){
  Y[i] ~ dbern(q[i])
  logit(q[i]) <- inprod(X[i,],beta[])
}

for(j in 1:p){beta[j] ~ dnorm(0,0.001)}
}")

# Compile model
logistic_model <- jags.model(model_string, data = dat_list, 
                             n.chains = 3, quiet = TRUE)

# Burn in samples
update(logistic_model, 2000, progress.bar="none")

# Posterior sampling
post_samples <- coda.samples(logistic_model, variable.names = "beta", 
                             n.iter = 50000, thin = 30, 
                             progress.bar="none")
```

---

# Results

.large[
```{r echo=FALSE, results = "asis"}
post_sum <- tidyMCMC(post_samples, conf.int = TRUE)
colnames(post_sum)[2:5] <- c("Mean","SD", "2.5%", "97.5%")
post_sum[,1] <- colnames(X)
knitr::kable(post_sum, format = "html", digits = 2)
```

<br>

.content-box-yellow[
**Your turn**

Provide an interpretation of the `Treatment` coefficient in context.
]

]




---

# Puffin nesting sites

.large[
Four variables where considered in trying to describe the nesting frequency of the common puffin in a 3m &#215; 3m grid of plots.


Variable | Description
---------|--------------------
`nests`  | number of nests (burrows) per 9 m<sup>2</sup> quadrat
`grass`    | grass cover percentage
`soil`     | mean soil depth in cm
`angle`    | angle of slope in degrees
`distance` | distance from cliff edge in meters
]

<br>
<br>

.footnotesize[Source: Nettleship, D. N. (1972). Breeding success of the Common Puffin (Fratercula arctica L.) on different habitats at Great Island, Newfoundland. *Ecological Monographs*, 42(2), 239-268.]


---
class: inverse

# Your turn

.Large[
1. What distribution should we use for the response variable, `nests`?

2. How can you transform a linear combination of the predictor to take on values in $(0, \infty)$?

$$\lambda_i = \beta_0 + \beta_1 {\tt soil}_i + \beta_2 {\tt distance}_i$$
]

---

# Data preparation

.large[Load and manipulate data]

```{r}
puffin <- read.table(
  "http://www.markirwin.net/stat149/Data/puffin.txt", header = TRUE
)

colnames(puffin)[1] <- "nests"
puffin <- subset(puffin, nests > 0)
```

.large[Data list for JAGS]

```{r}
X <- cbind(intercept = 1, puffin[,c("soil", "distance")])

dat_list2 <- list(Y = puffin$nests, X = X, 
                 n = nrow(puffin), p = ncol(X))
```

---

# Model fitting

```{r cache=TRUE, message=FALSE}
# Logistic regression specification
puffin_string <- textConnection("model{
for(i in 1:n){
  Y[i] ~ dpois(lambda[i])
  log(lambda[i]) <- inprod(X[i,], beta[])
}

for(j in 1:p){beta[j] ~ dnorm(0,0.001)}
}")

# Compile model
puffin_model <- jags.model(puffin_string, data = dat_list2, 
                           n.chains = 3, quiet = TRUE)

# Burn in samples
update(puffin_model, 2000, progress.bar="none")

# Posterior sampling
puffin_samples <- coda.samples(puffin_model, variable.names = "beta", 
                               n.iter = 100000, thin = 30, 
                               progress.bar="none")
```

---

# Results

.large[
```{r echo=FALSE, results = "asis"}
puffin_sum <- tidyMCMC(puffin_samples, conf.int = TRUE)
colnames(puffin_sum)[2:5] <- c("Mean","SD", "2.5%", "97.5%")
puffin_sum[,1] <- colnames(X)
knitr::kable(puffin_sum, format = "html", digits = 4)
```

<br>

.content-box-yellow[
**Your turn**

Provide an interpretation of the `distance` coefficient in context.
]

]